Efficiently Handling Guards in the low level design of RPython's tracing JIT

### 2012-07-16 talk/vmil2012/paper.tex
update year
show comments and add a command for my comments
set font size to 10pt
write more about the low-level aspects of guards

\documentclass{sigplanconf}

@@

\newboolean{showcomments}
\setboolean{showcomments}{false}
\ifthenelse{\boolean{showcomments}}
  {\newcommand{\nb}[2]{
    \fbox{\bfseries\sffamily\scriptsize#1}
    {\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}
   }
   \newcommand{\version}{\emph{\scriptsize$-$Id: main.tex 19055 2008-06-05 11:20:31Z cfbolz $-$}}
  }
  {\newcommand{\nb}[2]{}
   \newcommand{\version}{}
  }

\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\toon[1]{\nb{TOON}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand\fijal[1]{\nb{FIJAL}{#1}}
\newcommand\pedronis[1]{\nb{PEDRONIS}{#1}}
\newcommand{\commentout}[1]{}

@@

\conferenceinfo{VMIL'11}{}
\CopyrightYear{2012}
\crdata{}

@@

The contributions of this paper are:
\begin{itemize}
 \item
\end{itemize}

@@

The RPython language and the PyPy Project were started in 2002 with the goal of
creating a python interpreter written in a High level language, allowing easy
language experimentation and extension. PyPy is now a fully compatible
alternative implementation of the Python language, xxx mention speed. The
Implementation takes advantage of the language features provided by RPython
such as the provided tracing just-in-time compiler described below.

RPython, the language and the toolset originally developed to implement the
Python interpreter have developed into a general environment for experimenting
and developing fast and maintainable dynamic language implementations. xxx Mention
the different language impls.

@@

* Low level handling of guards
   * Fast guard checks v/s memory usage
   * memory efficient encoding of low level resume data
   * fast checks for guard conditions
   * slow bail out

% section Guards in the Backend (end)

>>>>

\documentclass[10pt]{sigplanconf}

@@

\newboolean{showcomments}
\setboolean{showcomments}{true}
\ifthenelse{\boolean{showcomments}}
  {\newcommand{\nb}[2]{
    \fbox{\bfseries\sffamily\scriptsize#1}
    {\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}
   }
   \newcommand{\version}{\emph{\scriptsize$-$Id: main.tex 19055 2008-06-05 11:20:31Z cfbolz $-$}}
  }
  {\newcommand{\nb}[2]{}
   \newcommand{\version}{}
  }

\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\toon[1]{\nb{TOON}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand\fijal[1]{\nb{FIJAL}{#1}}
\newcommand\pedronis[1]{\nb{PEDRONIS}{#1}}
\newcommand\bivab[1]{\nb{DAVID}{#1}}
\newcommand{\commentout}[1]{}

@@

\conferenceinfo{VMIL'12}{}
\CopyrightYear{2012}
\crdata{}

@@

In this paper we describe and analyze how deoptimization works in the context
of tracing just-in-time compilers. What instructions are used in the
intermediate and low-level representation of the JIT instructions and how these
are implemented.

Although there are several publications about tracing jut-in-time compilers, to
our knowledge, there are none that describe the use and implementaiton of
guards in this context. With the following contributions we aim to sched some
light (to much?) on this topic.
The contributions of this paper are:
\begin{itemize}
 \item
\end{itemize}

@@

The RPython language and the PyPy Project were started in 2002 with the goal of
creating a python interpreter written in a High level language, allowing easy
language experimentation and extension. PyPy is now a fully compatible
alternative implementation of the Python language\bivab{mention speed}. The
Implementation takes advantage of the language features provided by RPython
such as the provided tracing just-in-time compiler described below.

RPython, the language and the toolset originally developed to implement the
Python interpreter have developed into a general environment for experimenting
and developing fast and maintainable dynamic language implementations.
\bivab{Mention the different language impls}

@@

Code generation consists of two passes over the lists of instructions, a
backwards pass to calculate live ranges of IR-level variables and a forward one
to emit the instructions. During the forward pass IR-level variables are
assigned to registers and stack locations by the register allocator according to
the requirements of the to be emitted instructions. Eviction/spilling is
performed based on the live range information collected in the first pass. Each
IR instruction is transformed into one or more machine level instructions that
implement the required semantics.  Guards instructions are transformed into
fast checks at the machine code level that verify the corresponding condition.
In cases the value being checked by the guard is not used anywhere else the
guard and the operation producing the value can merged, reducing even more the
overhead of the guard.

Each guard in the IR has attached to it a list of the IR-variables required to
rebuild the execution state in case the trace is left through the side-exit
corresponding to the guard. When a guard is compiled, additionally to the
condition check two things are generated/compiled. First a special
data structure is created that encodes the information provided by the register
allocator about where the values corresponding to each IR-variable required by
the guard will be stored when execution reaches the code emitted for the
corresponding guard. \bivab{go into more detail here?!} This encoding needs to
be as compact as possible to maintain an acceptable memory profile.

\bivab{example goes here}

Second a trampoline method stub is generated. Guards are usually implemented as
a conditional jump to this stub, jumping to in case the guard condition check
does not hold and a side-exit should be taken. This stub loads the pointer to
the encoding of the locations mentioned above, preserves the execution state
(stack and registers) and then jumps to generic bail-out handler that is used
to leave the compiled trace if case of a guard failure.

Using the encoded location information the bail-out handler reads from the
saved execution state the values that the IR-variables had  at the time of the
guard failure and stores them in a location that can be read by the fronted.
After saving the information the control is passed to the frontend signaling
which guard failed so the frontend can read the information passed and restore
the state corresponding to the point in the program.

As in previous sections the underlying idea for the design of guards is to have
a fast on trace profile and a potentially slow one in the bail-out case where
the execution takes one of the side exits due to a guard failure. At the same
time the data stored in the backend needed to rebuild the state should be be
as compact as possible to reduce the memory overhead produced by the large
number of guards\bivab{back this}.

%* Low level handling of guards
%   * Fast guard checks v/s memory usage
%   * memory efficient encoding of low level resume data
%   * fast checks for guard conditions
%   * slow bail out
%
% section Guards in the Backend (end)
